---
title: "Stability Assignment"
author: "Erica Bishop"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(sensitivity)
library(tidyverse)
library(deSolve)
library(ggpubr)
```
### Defining stability

For this model, I'm defining stability as constant (or near constant) predator and prey populations for two decades. In the explorations below, I can mostly achieve stability within the 500 days, but I've also expanded and contracted the time window some. 

```{r source functon}
#source function
source("hunting_mod.R")

# set parameter list
pars = data.frame(rprey=0.95,
                  alpha=0.01,
                  eff=0.6,
                  pmort=0.4,
                  K=2000,
                  h = 0.2, #hunting rate
                  preypop = 200) #minimum prey population must be 400 for hunting to take effect
                  #prey pop only used for first graph - I think its more effective / realistic to use the prey parameter instead


```


Explore how different hunting levels and different minimum prey populations (before hunting is allowed) are likely to effect the stability of the populations of both predator and prey. Use this exploration to recommend a hunting target that will be sustainable (e.g leave you with a stable prey and predator population)

You can assume the following rprey=0.95, alpha=0.01, eff=0.6,pmort=0.4, K=2000,

# Exploration

The graphs and code below are where I tested many parameter combinations. I didn't keep everything I tried, so the final two graphs at the end of this docuent will be the most useful to look at. 

```{r test function}

# try initial conditions that will be stable
preyi = with(pars, pmort/(eff*alpha)) 
predi = with(pars, rprey/alpha*(1-preyi/K)) 

# times when you want to evaluate
days = seq(from=1,to=500)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p1 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line()

p1


```

This initial test does stabilize (yay!) - but I would like to see if I can get it to stabilize at a higher population, especially becuase as it stands hunting is only occuring when prey > 400. 

Next I will adjust my function to allow for hunting when prey is as low as 100 to see how this changes - I also fized this new function to account for the initial prey population, not a separate parameter (as in the function used above)

```{r}

#exploring different hunting thresholds in the function, so creating a new file here to not remove old results yet

#source new function with lower threshold
source("lotvmod_hunt.R")

#using same parameters as above
# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p2 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line()

p2


```
Setting the hunting threshold with the prey population de-stabilizes it a lot! 

Try increasing the initial conditions: 

```{r}

# times when you want to evaluate
days = seq(from=1,to=500)

# increasing initial populations
currpop=c(prey=100, pred=10)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p3 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line()

p3


```
Interestingly that decreased populations even more

Now I'll try going back to a 1 to 1 initial condition and messing with some of the other parameters

```{r}

# set new parameter list
pars = data.frame(rprey=0.95,
                  alpha=0.01,
                  eff=0.6,
                  pmort=0.4,
                  K=2000,
                  h = 0.02) #decreasing the hunting rate significantly

# time
days = seq(from=1,to=500)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p4 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line()

p4


```
The lower hunting rate didn't actually increase the populatin much at all, so instead I will leave a higher hunting rate and change the other parameters around 

```{r}

# set new parameter list
pars = data.frame(rprey=0.98,
                  alpha=0.01,
                  eff=0.3,
                  pmort=0.2,
                  K=2000,
                  h = 0.9) 

# time
days = seq(from=1,to=500)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p5 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 500))

p5


```
Playing around with the parameters a lot - having the hunting rate so high (as above) means that the prey population is pretty much only controlled by hunting and the predator population can grow over time - however they both dissappear after 25 days???
```{r}

# set new parameter list
pars = data.frame(rprey=1.2,
                  alpha=0.01,
                  eff=0.3,
                  pmort=0.2,
                  K=2000,
                  h = 0.2) 

# time
days = seq(from=1,to=500)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p6 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 500))

p6



```
```{r}

source("lotvmod_hunt2.R") #raised hunting threshold to try to not have them all die? - raised and lowered


# set new parameter list
pars = data.frame(rprey=1.8,
                  alpha=0.01,
                  eff=0.3,
                  pmort=0.3,
                  K=2000,
                  h = 0.9) 

# time
#hunting has a more noticeable effect before pop really stabilizes
days = seq(from=1,to=200)

# first see what happens when we start with 1 of each
currpop=c(prey=1000, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p7 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() 

p7



```
The exploration above might be a bit hard to follow (sorry!) But I tried to document some of the things I was testing. I found that I really had to decrease the prey popiulation threshold for hunting to see major effects in the graphs. I also had trouble getting predator and prey populations to stabilize at high populatons, they seemed happiest around 100ish (or always below 500) - so the carrying capacity never really took effect in this model. 

The if / else statement I originally had in the function was also causing issues where both populations could entirely die off - so I decidied to leave the simple model with hunting always incorporated

# Testing stability

As defined above, I'm looking for populations to be stable for two decades into the future. Below I compare how a low hunting rate and a high hunting rate affect population stability over 
```{r}
# using original parameters to really focus on hunting effect
pars = data.frame(rprey=0.95,
                  alpha=0.01,
                  eff=0.6,
                  pmort=0.4,
                  K=2000,
                  h = 0.9)

# time - 20 years
days = seq(from=1,to=7305)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p8 = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 7305))

p8


```


```{r}
source("lotvmod_hunt2.R")

# using original parameters to really focus on hunting effect
pars = data.frame(rprey=0.95,
                  alpha=0.01,
                  eff=0.6,
                  pmort=0.4,
                  K=2000,
                  h = 0.001) #starting with a really low hunting rate, jsut 0.1%

# time - 20 years
days = seq(from=1,to=7305)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p_lowhunt = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 7305))

```


```{r}

# using original parameters to really focus on hunting effect
pars = data.frame(rprey=0.95,
                  alpha=0.01,
                  eff=0.6,
                  pmort=0.4,
                  K=2000,
                  h = 0.99) #really high hunting rate

# time - 20 years
days = seq(from=1,to=7305)

# first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)

# run differential equation solver
res = ode(func=lotvmod_hunt, y=currpop, times=days, parms=pars)

# extract the results
res_initial = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p_highhunt = ggplot(res_initial,
          aes(time, pop, col=animal)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 7305))


```

```{r}

ggarrange(p_lowhunt, p_highhunt)


```

### Recommendation for sustainable hunting

Based on my exploration here, hunting rates can actually remain someone high, up to 0.9 - with other parameters held constant - and the population will remain stable. However, at the 0.9 hunting rate (seen in p8 graph) - the prey population is really low even though its defined as "stable" in my definition. Therefore I would recommend a more nuanced definition of stability and a lower hunting rate that allows for a higher population. 

 